\addchap{Acknowledgements}

The last few years would not have been as interesting and enjoyable without the help and the company of certain special people.
Primarily, I am very grateful to Hans Peter Büchler for a very inspiring time and the chance to work on a lot of interesting projects.
I have learned a lot from you and I very much enjoy discussing physics with you.
Thank you for always supporting me and for giving me many valuable opportunities throughout my PhD.
I want to thank Tilman Pfau for being the first one to get me interested in this exciting field of physics.
You have supported me in many different ways throughout my years at the university and I am very grateful for that.
I am very thankful to Maria Daghofer for promptly accepting to review my thesis.

I have profited a lot from discussing and doing physics with many great people I have met during the last years.
Special thanks go to my fellow PhD students Nicolai Lang and Przemysław Bienias for uncounted hours of interesting and very helpful discussions.
Further thanks for invaluable discussions go to Jonathan Balewski, Sebastian Weber, Norman Yao, Alban Urvoy, Axel Griesmaier, Sebastian Huber, Thierry Lahaye, Antoine Browaeys and Kazimierz Rzążewski.
Particular thanks also goes to friends and colleagues I have met throughout the years: Adam Bühler, Thorsten Treffon, Krzysztof Pawłowski, Jens Honer, Stephan Humeniuk, Michael Hauber, Eslam Khalaf, Karim Bouadim and Anita Gaj.
It would not have been the same without you.
Thanks also go to Mikhail Lukin for giving me the opportunity to join his group in Harvard for two months and to Ildiko Poljak and Oliver Nagel for the organizational part.
A huge thank you goes to Andi, Basti, Fabi, Holger, Matze and Patrick for many memorable experiences throughout the last nine years.
Last, extraordinary thanks go to my family for always supporting me!
